---
title: "NiN analysis Innlandet"
format: html
author: Øystein Sporild Myran
editor: visual 
lang: no
---

GIS PORTFOLIO PROJECT:

NiN-nature types within simulated 10km x 10km planning area

Innlandet, Norway

# Introduction

This project was created as a way for me to further develop my GIS and spatial analysis skills, building upon what I learned during my time as a master's student in bioscience at the University of Oslo. The project demonstrates a basic spatial analysis workflow in R using "nature type" data based on the Norwegian Nature in Norway (NiN) classification system. The goal was simple: identify and summarize mapped nature types within a selected planning area in Innlandet county.

# Data

## Setup

```{r load-packages}
# Load packages
library(sf)
library(dplyr)
library(ggplot2)
library(ggspatial)
```

## Load data

The NiN nature type dataset was imported from an ESRI geodatabase (EPSG:25833) downloaded from Miljødirektoratet.

```{r load-data}
# Make path to data
gdb_path <- "~/GIS_data/NiN_data/naturtyper_innlandet.gdb"

# Read nature type data
natur <- st_read(
  gdb_path,
  layer = "naturtyper_nin_omr"
)
```

# Methods

First, the coordinate reference system (CRS) and spatial extent of the dataset were inspected to ensure that all layers used compatible spatial projections. The center point of the dataset extent was then identified and used to define a smaller planning area for analysis and visualization.

A rectangular planning area was created as an `sf` polygon object around the selected center coordinates. Spatial intersection analysis was then performed using `st_intersection()` to clip nature type polygons to the planning area, only retaining the portions of mapped nature types located within the analysis boundary.

The area of each resulting polygon was calculated using `st_area()` and converted from square meters to hectares. Areas were then summarized by NiN nature type using functions from the `dplyr` package.

In order to create a easily interpretable map, only the nature types covering the largest geographical area were used in visualization.

Finally, the results were visualized using `ggplot2` and `ggspatial` to create a clean map suitable for a portfolio, showing the spatial distribution of mapped nature types within the selected planning area.

## Inspect data

```{r inspect-data, include=FALSE}
# Retrieve coordinate reference system
st_crs(natur)
# Return bounding of the data
st_bbox(natur)
```

## Find center of data set

```{r find-center}
sentrum <- st_centroid(st_union(natur))
coords <- st_coordinates(sentrum)

x_mid <- coords[1]
y_mid <- coords[2]
```

# Spatial analysis

## Create planning area

```{r create-planning-area}
# Create a simulated 10km x 10km planning area
plan_coords <- matrix(
  c(
    x_mid - 5000, y_mid - 5000,
    x_mid + 5000, y_mid - 5000,
    x_mid + 5000, y_mid + 5000,
    x_mid - 5000, y_mid + 5000,
    x_mid - 5000, y_mid - 5000
  ),
  ncol = 2,
  byrow = TRUE
)

plan_polygon <- st_polygon(list(plan_coords))

plan_sf <- st_sf(
  geometry = st_sfc(
    plan_polygon,
    crs = st_crs(natur)
  )
)
```

## Spatial intersection

```{r spatial-intersection}
#| warning: false
natur_klippet <- st_intersection(natur, plan_sf)
```

## Area calculations

```{r area-calculations}
natur_klippet$Areal_ha <-
  as.numeric(st_area(natur_klippet)) / 10000

# Finding sum of mapped areas i hectare
total_areal_kartlagt <- sum(natur_klippet$Areal_ha)

print(total_areal_kartlagt)

# Calculating % of total planning area mapped
andel_kartlagt <- (total_areal_kartlagt / 10000) * 100

print(andel_kartlagt)
```

## Summarize nature types

```{r summarize, include=FALSE}
#| warning: false
natur_tabell <- st_drop_geometry(natur_klippet)

areal_per_type <- natur_tabell %>%
  group_by(naturtype) %>%
  summarise(
    Areal_ha = sum(Areal_ha, na.rm = TRUE)
  ) %>%
  arrange(desc(Areal_ha))
```

# Visualization

## Choosing largest nature types for visualization

```{r largest-nature-types}
top_types <- areal_per_type %>%
  slice_max(Areal_ha, n = 5) %>%
  pull(naturtype)

natur_klippet$naturgruppe <- ifelse(
  natur_klippet$naturtype %in% top_types,
  natur_klippet$naturtype,
  "Andre naturtyper"
)
```

## Plotting with ggplot2

```{r p_natur_klippet-plot}
# Bounding box around data
bbox <- st_bbox(natur_klippet)

# Plot map
p_natur_klippet <- ggplot() +

  # Nature types
  geom_sf(
    data = natur_klippet,
    aes(fill = naturgruppe),
    color = "grey30",
    linewidth = 0.1,
    alpha = 0.9
  ) +

  # Zoom to data
  coord_sf(
    xlim = c(bbox$xmin - 500, bbox$xmax + 500),
    ylim = c(bbox$ymin - 500, bbox$ymax + 500),
    expand = FALSE
  ) +
  
  # Natural color palette 
  scale_fill_brewer(
    palette = "PuOr",
    name = "Naturtype"
  ) +

  # North arrow
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    style = north_arrow_orienteering(),
    height = unit(0.8, "cm"),
    width = unit(0.8, "cm"),
    pad_x = unit(0.5, "cm")
  ) +

  # Scale
  annotation_scale(
    location = "bl",
    width_hint = 0.3
  ) +

  # Minimalistic theme
  theme_minimal(base_size = 12) +

  labs(
    title = "Kartlagte NiN-naturtyper",
    subtitle = "Simulert analyseomrade i Innlandet",
    caption = "Kilde: Miljodirektoratet - Natur i Norge (NiN) kartdata",
    fill = "Naturtype"
  ) +

  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    axis.title = element_blank(),
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    
    plot.caption = element_text(
    size = 8,
    hjust = 0,
    color = "grey40"
  ),
    legend.position = "right"
  )

# Print plot
p_natur_klippet

```

```{r save-as-PNG}
# Save map as PNG in figures-folder
ggsave(
  filename = "figures/portfolio_map_clean.png",
  plot = p_natur_klippet,
  width = 10,
  height = 7,
  dpi = 300
)
```

# Discussion

The spatial analysis identified a range of mapped nature types within the the selected planning area. Within the 10km $\times$ 10km planning area, mapped nature type polygons covered approximately 414 hectares, corresponding to about 4.1$\%$ of the total area. This reflects the selective and time consuming nature of ecological field mapping, where the standards and focus of the mapping procedure is set by the project client.

There are four types that stand out in terms of geographical coverage: *Kalkfattig og intermediaer fjellhei, leside og tundra* (347 ha), *Boreal hei* (36.3 ha), *Naturbeitemark* (16.3 ha) and *Hogreliggende og nordlig nedborsmyr* (12.1 ha). Together, these account for 411.7 hectares, corresponding to 99.4$\%$ of the total mapped nature type area.

This project provided practical experience working with spatial ecological data in R and served as a refresher on GIS workflows involving NiN datasets. In addition to the technical aspects of the spatial analysis, the project also highlighted several important considerations in cartographic visualization, including the selection of an appropriate planning area, decisions regarding which spatial information to include, and the importance of effective color palette design for communicating ecological patterns in a clear way.
